Relationships in the extremes and their influence on competition and coexistence

### 2021-09-17 paper/Paper.Rnw
comments for paragraph 4
changes made to Intro in meeting with Jasmin

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that make storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
%This is reasonable assumption for non-rare species, since 
%a better environment leads to greater numbers which leads to more intraspecific competition, but a rare species may be essentially unaffected
%by intraspecific competition. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor; dormant stages such as are in seed banks are an
example buffering mechanisms \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects are, in fact, quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, so ATAs 
between between E and C may mediate the strength of storage effects.
%***Old text below, kept for now, but should be pretty much replaced by the above
%Storage effects allow for species to fluctuate over time while escaping extinction, continously recovering from low densities. 
%In order for storage effects to contribute as a stabilizing mechanism to coexistence, there must first be covariation between competitive pressures and environmental fluctuations [EC covariance \citep{Chesson_1994, Barabas_2018}].
%Because the contributive value of storage effects relies on the relationship between environment and competition \citep{Chesson_1994, Barabas_2018}, it reasonably follows to suspect the contribution of storage effectsto be altered by an association between environment and competition that is asymmetric, i.e., ATAs between environment and competition.
%The reliance of storage effect's magnitude on the degree of relationship between environment and competition indicate that storage effects involve environment-coexistence associations, and hence there is potential for ATAs to influence storage effects. 
%In addition to EC covariance, functioning temporal storage effects also require that heterospecific populations respond differently to environmental fluctuations, and different life stages must also vary in responses \citep{}. 
%In a simple two species system, differences in environmental response allow for one species to grow to high densities in its favorable environemnt while the other is very low. %**JAS: should we introduce invader analysis or terminology here?
%Growth is stable for the high density species due to positive EC covariance -- high intraspecific competition limits its growth rate in a favorable environment. 
%Conversely, EC covariance is lesser for the low density species, leading to large fluctuations in population 
%growth rate in both directions. 
%Although, this may appear like a scenario for extiction, it can, in fact, be advantageous due to differences between life stages, buffering species against extinction by limiting negative growth rates, but not positive ones. 
%For example, insensitive adults or dormants seeds can still persist when in an unfavorable environment and overall growth rate is negative \citep{Chesson_2012}; environment and competiton do not effect the buffered life stages.   
%Thus, averaged over multiple generations, higher fluctuation in growth rate can lead to a positive low density growth rate and coexistence is possible \citep{Chesson_2000}.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events.

%***DAN: I think the above paragraphs are pretty solid. I am still working on the below.

%[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
%There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here  In one of the first empirical tests for storage effect and its requirements, \cite{Adler_2006} showed that climate variability - including extreme events - promoted the coexistence of grass species in Kansas prairie, specifically their ability to recover from low densities. Moreover, in studies on the invasibility of plants, the density and competitive balance of invasive Japanese knotweed increases as resource and thus climate variability increased.\citep{Parepa_2013}

%***DAN: Jasmin, there are a lot of disjointed sentences in the below paragraph, at least they appear disjointed to me.
%Can you please say in a comment right below this one what the topic sentence is? As usual, if you are having trouble 
%determining what the topic sentence it, it may mean you don't have one, and you need to determine what you're doing 
%with this paragraph and make a topic sentence. 

@@

@@

@@

In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
%**JAS: I intended for the first sentence to be the topic sentence 
% a better one:
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
Extreme events can have important ecological impacts for
%***DAN: I don't know what "controphic" means and when I google it I get nothing. Perhaps you intended a different word?
% can cut out 
competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

To demonstrate plausibility of ATA altering coexistence between species, we carried out long time scale simulations using a simple two-species lottery model, and below (in the next paragraph), we describe a hypothetical example. Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. We simulated two different communites up to \Sexpr{M} time steps to hopefully make it clear to the reader that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog}). The only distinction between the two simulations is that one depends on asymmterically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverablility when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

To contextualize idea of how ATA in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}). The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. The result is a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated, and, following our simulated example (Figure \ref{fig:pedagog}), the species persist in coexistence. However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the higher values or right-tail associated. Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.

@@

%Ideas while driving to pick up Max:
%1) In this paper with Jasmin you need to cite the two papers you just got from the professor at University of Massachusetts, Lowell. I mean the papers about changes in the special extent of heat waves: https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2020GL087097 and https://iopscience.iop.org/article/10.1088/1748-9326/ab4b41
%2) Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model. Probably real environmental variables are actually related to vital rates in a non-linear and probably often threshold-like way. Threshold relationships can produce tail association, as we showed elsewhere. Changes in the means of true environmental variables may change where those distributions of environmental variables are compared to thresholds, which may change the nature of tail associations between the effects of those environmental variables.

>>>>

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that make storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
%This is reasonable assumption for non-rare species, since 
%a better environment leads to greater numbers which leads to more intraspecific competition, but a rare species may be essentially unaffected
%by intraspecific competition. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor; dormant stages such as are in seed banks are an
example buffering mechanisms \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects are, in fact, quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, so ATAs 
between between E and C may mediate the strength of storage effects.

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
%**JAS: I intended for the first sentence to be the topic sentence 
% a better one:
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
Extreme events can have important ecological impacts for
competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). 
%***DAN: Leave the following for the next para
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

%***DAN:
%1) Topic sentence: To make the above ideas more concrete, we review a relevant example from the literature that may be
%influenced as described above. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
To contextualize idea of how ATA in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}). The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. The result is a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated, and, following our simulated example (Figure \ref{fig:pedagog}), the species persist in coexistence. However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the higher values or right-tail associated. Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

@@

% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??
%***DAN: Possible topic sentence like thing after talking to Jasmin on 9/17: 
%Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied,
%the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, 
%have been studied in several systems. 
%***DAN: Do it like you have it but shorter.

@@

%Jasmin tried to split this para up into effects on coexistence of changes in mean, starting here... 

@@

%...and variance...
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. 
%...and extreme events (though this blends together with changes in variance)
Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.
%DAN: So what jasmin was trying to do was summarize a lot of lit on effects of climate change on coexistence, in these three parts
%(means, varinace, and extreme events).

%Move up the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model.  We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
%Lotetry model intro
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
lottery model, and below (in the next paragraph), we describe a hypothetical example. Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. We simulated two different communites up to \Sexpr{M} time steps to hopefully make it clear to the reader that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog}). The only distinction between the two simulations is that one depends on asymmterically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverablility when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

%Goal para
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.

@@

%Ideas while driving to pick up Max:
%1) In this paper with Jasmin you need to cite the two papers you just got from the professor at University of Massachusetts, Lowell. I mean the papers about changes in the special extent of heat waves: https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2020GL087097 and https://iopscience.iop.org/article/10.1088/1748-9326/ab4b41
%The second one is the one to cite
%2) Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model. Probably real environmental variables are actually related to vital rates in a non-linear and probably often threshold-like way. Threshold relationships can produce tail association, as we showed elsewhere. Changes in the means of true environmental variables may change where those distributions of environmental variables are compared to thresholds, which may change the nature of tail associations between the effects of those environmental variables.
